---
title: "Dataset Visualization"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Read Data
```{r}
dataset <- read.table(file="datasets/achilles/Achilles_v3.3.8_rawreadcounts.txt",header=T,sep='\t',as.is=T)
```

### Add gene column
```{r}
mappings <- read.table(file="datasets/achilles/Achilles_v3.3.8_sgRNA_mappings.txt",header=T,sep='\t',as.is=T)
underscores <- regexpr("_", mappings$Guide)
mappings$Name <- substring(mappings$Guide, 1, underscores-1)
mappings$Gene <- substring(mappings$Guide, underscores+1)
counts <- merge(dataset, unique(mappings[,c("Name","Gene")]))
```

### Add category column:
```{r}
library(openxlsx)
ranking_hart <- openxlsx::read.xlsx("datasets/hart/msb0010-0733-sd1.xlsx", sheet=1)
essentials <- na.omit(ranking_hart$`ConstitutiveCoreEssentials(CCE)`)
nonessentials <- na.omit(ranking_hart$`Nonessential.Genes.(NE)`)
counts <- counts[-grep("NO_CURRENT_A|NO_CURRENT_C|NO_CURRENT_G|NO_CURRENT_T",counts$Gene),]
counts$Category <- "Other"
counts$Category[grep("NO_CURRENT", counts$Gene)] <- "Nontargeting"
counts$Category[counts$Gene %in% essentials] <- "Essential"
counts$Category[counts$Gene %in% nonessentials] <- "Nonessential"

# Rearrange the columns to have the information at the beginning:
counts$Description <- NULL
info <- which(names(counts) %in% c("Name","Gene","Category"))
counts <- cbind(counts[,info], counts[,-info])
info <- which(names(counts) %in% c("Name","Gene","Category"))

# Some of the rows are repeated (because the same spacer can correspond to multiple genes), so we need to remove them when doing calculations:
repetitions <- table(counts$Name)
repetitions <- repetitions[repetitions>1]

# Remove repeated rows:
counts <- subset(counts, !duplicated(counts$Name)) 
repetitions <- table(counts$Name)
repetitions <- repetitions[repetitions>1]
write.table(counts,"Achilles_june27.txt",quote=F,sep='\t',row.names=F,col.names=T)
```

### Load data
```{r}
counts <- read.table("Achilles_june27.txt",header=T)
info <- c(1,2,3)
```

### Category distribution of sgRNA pairs
```{r}
library(ggplot2)
my_data <- data.frame(table(counts$Category))
names(my_data) <- c("Category", "Frequency")
p1 <- ggplot(my_data, aes(x=factor(1), y=Frequency, fill=Category)) + geom_bar(stat="identity") + 
  coord_polar(theta = "y") + xlab("") + 
  theme(axis.text.y=element_blank(),axis.ticks.y=element_blank()) + ylab("")
ggsave("pie_chart.png",p1)
```

### Plot Counts vs Sample
```{r}
library(ggplot2)
library(reshape2)
my_data <- split(counts[,-info], counts$Category)
my_data <- do.call(rbind, lapply(my_data, colSums))
my_data <- melt(my_data, value.name="Sum", varnames=c("Category","Sample"))
b1 <- ggplot(my_data, aes(x = Sample, y = Sum, fill=Category)) + geom_bar(stat = "identity") +
  theme(axis.text.x=element_text(angle=90,hjust=1), legend.position="bottom") + xlab("")
ggsave("bar_chart.png",b1)
```

### Number of sgRNAs with zero counts
```{r}
library(ggplot2)
library(reshape2)
my_data <- split(counts[,-info], counts$Category)
my_data <- do.call(rbind, lapply(my_data, function(x) apply(x,2,function(y) sum(y==0))))
my_data <- melt(my_data, value.name="Zeroes", varnames=c("Category","Sample"))
ggplot(my_data, aes(x = Sample, y = Zeroes, fill=Category)) + geom_bar(stat = "identity") +
  theme(axis.text.x=element_text(angle=90,hjust=1), legend.position="bottom") + xlab("")
```

### Make a table of sgRNAs ranked by counts
```{r,fig.height=4}
library(gridExtra)
tt1 <- ttheme_minimal(base_size=8,
  core=list(fg_params=list(fontface=3),bg_params=list(fill=c("lightblue","lightgreen"),alpha=0.5,col=NA)),
  colhead=list(fg_params=list(col="white",fontface=4L),bg_params=list(fill="darkblue")),
  rowhead=list(fg_params=list(col="orange", fontface=3L),bg_params=list(fill="white")))
tt2 <- ttheme_minimal(base_size=7,
  core=list(fg_params=list(fontface=3),bg_params=list(fill=c("pink","chocolate"),alpha=0.5,col=NA)),
  colhead=list(fg_params=list(col="white",fontface=4L),bg_params=list(fill="darkred")),
  rowhead=list(fg_params=list(col="orange", fontface=3L),bg_params=list(fill="white")))
p1 <- tableGrob(counts[order(counts[,4],decreasing=T)[1:30],c(1:3,4)],theme=tt1,rows=NULL)
grid.arrange(p1,nrow=1)
library(knitr)
kable(counts[order(counts[,4],decreasing=T)[1:30],c(1:3,4)])
```

### Normalize by nonspecific genes and get log2 transform
```{r}
logged <- normed <- counts
normed[,-info] <- apply(counts[,-info], 2, function(x) x/median(x[counts$Category=="Nontargeting"]))
logged[,-info] <- log2(normed[,-info] + 0.05)
```

### Principal Components of the Samples: 2nd vs 1st
```{r,fig.width=10,fig.height=10}
library(ggrepel)
pca_sample <- prcomp(t(logged[,-info]),scale=F)
my_data <- data.frame(Z1=pca_sample$x[,1],Z2=pca_sample$x[,2],Sample=names(counts[,-info]))
ggplot(my_data, aes(Z1,Z2,label=Sample)) + geom_text_repel() + geom_point() + 
  xlab("First Principal Component for Samples") + ylab("Second Principal Component")
```

## Make a heatmap of euclidean distances between samples using first 3 principal components
```{r,fig.width=10,fig.height=10}
library(gplots)
my_data <- as.matrix(dist(pca_sample$x[,1:3], method="euclidean"))
heatmap.2(my_data,margins=c(11,11),density="density",tracecol="black",trace="none")
```

### Proportion of Variance Explained
```{r}
library(gridExtra)
pve_sample <- 100*pca_sample$sdev^2/sum(pca_sample$sdev^2)
my_data=data.frame(PC=seq(1:length(pve_sample)),pve_sample=pve_sample)
p1 <- ggplot(my_data,aes(PC,pve_sample)) + geom_line(col="blue") + geom_point(col="darkblue",shape="O",size=5) + xlab("Principal Component") + ylab("Proportion of Variance Explained")
my_data=data.frame(PC=seq(1:length(pve_sample)),Cpve_sample=cumsum(pve_sample))
p2 <- ggplot(my_data,aes(PC,Cpve_sample)) + geom_line(col="blue") + geom_point(col="darkblue",shape="O",size=5) + xlab("Principal Component") + ylab("Cumulative PVE")
grid.arrange(p1,p2,nrow=2)
```

### Hierarchical clustering of samples using the 3 first principal components
```{r,fig.width=10,fig.height=10}
clustered <- hclust(dist(pca_sample$x[,1:3]))
plot(clustered,labels=names(logged[,-info]),xlab="",sub="",ylab="")
```


